---
title: "Code Book to Accompany run_analysis.R script"
author: "Juha Ronkainen"
date: "17/05/2018"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, results=FALSE}
#Install packages, if needed
#The first block of code just makes sure, you have the downloader package installed, and if not it proceeds to install it. Note that the reusults of this code chunk are deliberately hidden, as it is just so boring and takes half of the document. 
list.of.packages <- c("downloader", "ggplot2", "timeDate", "gridExtra", "plyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(downloader)
library(ggplot2)
library(timeDate)
library(gridExtra)
library(plyr)
```
#Code Book to accompany run_analysis.R script
The aim of this Code Book is to explain how to get the data, 
what are the variables and what transformations has been done in order
to pull out the results. 

The data is available online:
```{r}
destfile="./Fprojectfiles.zip"
if (!file.exists(destfile)){
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip", destfile, method="curl")
unzData<-unzip("Fprojectfiles.zip", exdir = "./data/")
}
```
All text files included in the dataset have been transformed into data frames. It is a long and cumbersome bulk script, which you can check out in the run_analysis.R script. 

```{r echo=FALSE, results=FALSE}
csv_files_data <- list.files(path = "./data/UCI HAR dataset/", pattern = "*.txt", full.names=TRUE)
list2env(setNames(lapply(csv_files_data, read.csv), basename(tools::file_path_sans_ext(csv_files_data))), globalenv())

csv_files_data_test <- list.files(path = "./data/UCI HAR dataset/test", pattern = "*.txt", full.names=TRUE)
list2env(setNames(lapply(csv_files_data_test, read.csv), basename(tools::file_path_sans_ext(csv_files_data_test))), globalenv())

csv_files_data_train <- list.files(path = "./data/UCI HAR dataset/train", pattern = "*.txt", full.names=TRUE)
list2env(setNames(lapply(csv_files_data_train, read.csv), basename(tools::file_path_sans_ext(csv_files_data_train))), globalenv())
```

Here are the original txt files and the data frames are named accordingly as you can see in the example for features: 
```{r echo=FALSE, results=TRUE}
csv_files_data
csv_files_data_train
csv_files_data_train
head(features)
```

However, two main data sets are broken and cannot be parsed into data frames without some pain. We transform X_train.txt and X_test.txt into data frames by detecting the fixed widht of the columns. The amount of columns is deducted from the file features, as it has 561 rows.
```{r echo=FALSE, results=FALSE}
X_train<-read.fwf("./data/UCI HAR dataset/train/X_train.txt", widths=c(rep(16, 561)))
X_test<-read.fwf("./data/UCI HAR dataset/test/X_test.txt", widths=c(rep(16, 561)))

#Get column names for X_train and X_test
features <- read.table("./data/UCI HAR dataset//features.txt", sep=" ")
```

We removed the non needed row number from the features dataset and name the columns of our important X_test and X_train datasets. The names of these columns are detailed in the accompanying document features_info.txt
```{r echo=TRUE, results=FALSE}
features<-features[-1]

#Naming the columns of X_train and X_test
colnames(X_train)<-features[,1]
colnames(X_test)<-features[,1]
```

Then we manually (!) extracted only the mean and standard deviation records (columns) from the data frames. 

```{r echo=FALSE, results=FALSE}
x_train<-X_train[c(1:6, 41:46, 81:86, 121:126, 161:166, 201, 202, 214,215, 227, 228, 240, 241, 253,254, 266:271, 345, 350, 373,374,375, 424:429, 452:454, 503,504, 513, 516,517, 529, 530, 539, 542,543, 555:561)]
x_test<-X_test[c(1:6, 41:46, 81:86, 121:126, 161:166, 201, 202, 214,215, 227, 228, 240, 241, 253,254, 266:271, 345, 350, 373,374,375, 424:429, 452:454, 503,504, 513, 516,517, 529, 530, 539, 542,543, 555:561)]
```
Then we added columns on users (subject_train) and activity (y_test). Furthermore, we replaced activity codes (1-6) with more descriptive names as given in activity_labels.txt file. 

```{r echo=FALSE, results=FALSE}
x_train$subject<-subject_train[,1]
subject_test<-rbind(subject_test, 24)
x_test$subject<-subject_test[,1]
x_train$activity<-y_train[,1]

#x_test has a one missing value and it has been replaced 
#by the value just before it (assuming it is the same person)
y_test<-rbind(y_test, 2)
x_test$activity<-y_test[,1]

#To give the activities a descriptive names, we must create a small function
transform<-function(x){
        if(x==1) y<-"walking" else
        if(x==2) y<-"walking_upstairs" else
        if(x==3) y <-"walking_downstairs" else 
        if(x==4) y<-"sitting" else
        if(x==5) y<-"standing" else
        if(x==6) y<-"laying" else y<-NA
return(y)
}

#And let's now give activities descriptive names in data frams x_test and x_train
x_train$activityDescr<-lapply(x_train$activity, transform)
x_test$activityDescr<-lapply(x_test$activity, transform)
```

Finally we merged our two precious data frames x_train and x_test into a massive clean data frame, transformed that into a matrix and saved a csv file (x_trainAndtest.csv) for future generations to enjoy. 

```{r echo=FALSE, results=FALSE}
#And now, let's merge the data into one massive (but clean) dataset
x_trainAndtest<-rbind(x_train, x_test)

#And let's save that into a csv file
x_trainAndtestMatrix<-as.matrix(x_trainAndtest)
write.table(x_trainAndtestMatrix, file="x_trainAndtest.txt", row.name=FALSE)
```
Resulting matrix is sizable: 
```{r echo=FALSE, results=TRUE}
dim(x_trainAndtest)
```
and it looks like this:
```{r}
head(x_trainAndtest)
```

And finally, we created a second, independent tidy 
data set with the average of each variable for each activity and each subject. Also this was transformed into a matrix and saved as a csv file (x_averagesSubjectActivity.csv). 

```{r echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
#This is pretty easy with aggregate function
x_averagesSubjectActivity<-aggregate(x_trainAndtest, by=list(x_trainAndtest$activity, x_trainAndtest$subject), mean)
#We just add back the activity descriptions
x_averagesSubjectActivity$activityDescr<-lapply(x_averagesSubjectActivity$activity, transform)
#and write a new clean tidy data set into csv file
#And let's save that into a csv file
x_averagesSubjectActivity<-as.matrix(x_averagesSubjectActivity)
write.table(x_averagesSubjectActivity, file="x_avereagesSubjectActivity.txt", row.name=FALSE)
```

This second clean and neat data set is much more compact:
```{r echo=FALSE, results=TRUE}
dim(x_averagesSubjectActivity)
```
and it looks like this
```{r}
head(x_averagesSubjectActivity)
```